Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-27 manu-draft.qmd
methods draft done

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and the likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observation in the EPC data. To quantify daily counts each year, a continuous prediction of temperature or stress over time was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a]. Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).

Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the GAMs. This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.

All models excluded Lower Tampa Bay because of minimal seagrass change over time. Evaluated total fo, percent change between years, and

GLMs of percent change each year by Sal, Temp, with year and bay segment interactions. Attempted different windows of time from approximate seagrass sample date, 30 days to one year. The best model was one that included an effect of Salinity and an interaction with bay segment. Hillsborough Bay showed a strong response of increasing likelihood of seagrass loss with number of days when salinity was below the threshold.

Then GAMs were evaluated looking at total by Salinity, with interaction of year and bay segment. Also created a model looking at percent change vs Temp and year/bay segment interaction.

Show plots of both the glm and gams.

>>>>

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and the likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year [@wood17]. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a]. Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).

Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the daily GAM predictions. This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt. The number of days when both temperature was above and salinity was below the thresholds was also estimated as a combined potential stress measure. Stressor metrics were further aggregated across stations in each bay segment using a mixed-effects regression model where the annual stressor counts for stations in a bay segment were fit against year (1975 to 2022) using a random intercept for station [@zuur2009a; @bates15]. This produced an overall assessment of how the stressor metrics have changed over time by bay segment, including counts for temperature, salinity, and both relative to the thresholds.

For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above. Transect surveys are typically done in the fall at the end of the growing season. The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to various response measures at the bay segments-scale from the transect data, including frequency occurrence each year (all species), percent change in frequency occurrence between years, and increase/decrease in frequency occurrence between years as a binary (1/0) variable. Generalized linear regression models were used to evaluate each of the seagrass response variables relative to the stressor metrics. Interaction factors were also included for bay segment and time period (recovery pre-2016, decline post-2016). The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity. These models did not include stressor metrics, but rather used direct measurements of salinity and temperature as independent variables. Models of temperature or salinity to predict daily measurements are not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date. However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys. This allowed for more flexible models that are less limited by degrees of freedom. GAMs were used to evaluate seagrass percent cover for the FIM data using continuous temperature and salinity as independent variables. Interactions with factor variables for bay segment and time period as above were also included. The GAMs also evaluated interactions between salinity and temperature to determine if both factors had an additive effect on seagrass change. The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).
